---
title: "#TidyTuesday Look at Australian Birds"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(readxl)
library(cluster)
library(ggpubr)
library(factoextra)
library(ggrepel)
```

First things first, bring in the data. Here we go back to the file from the source (<https://doi.org/10.1371/journal.pone.0150899>) so that we can get the Birdbath IDs, which were lost in the TidyTuesday export

```{r data_in}
bird_baths <- read_excel("S1File.xlsx") %>%
  clean_names() %>%
  mutate(birdbath_number = row_number()) %>%
  relocate(birdbath_number) 

birds <- bird_baths  %>%
 select(!c(survey_year, urban_rural, bioregions, birdbath_number)) %>%
  t() %>%
  as_tibble(rownames = "bird_name", .name_repair = "universal") %>%
  rename(total = last_col()) %>%
  select(bird_name, total)

# transpose so we are looking at birds as observations and their appearance at given birdbaths as variables

bird_matrix <- bird_baths %>%
  drop_na(survey_year)%>%
  select(!c(survey_year, urban_rural, bioregions, birdbath_number)) %>%
  t()
```

Then we get the clusters based on which birdbaths they appear in. (Note the data considers a birdbath independently by season and year

Most of this clustering code is inspired by: <https://www.statmethods.net/advstats/cluster.html>

```{r}
# Determine number of clusters
wss <- (nrow(bird_matrix)-1)*sum(apply(bird_matrix,2,var))
for (i in 2:30) wss[i] <- sum(kmeans(bird_matrix,
   centers=i)$withinss)
plot(1:30, wss, type="b", xlab="Number of Clusters",
  ylab="Within groups sum of squares")

```

```{r}

# K-Means Cluster Analysis
fit <- kmeans(bird_matrix, 4) # 4 cluster solution seems to work well

# append cluster assignment
kmeans_out <- as.data.frame(fit$cluster) %>%
  as_tibble(rownames = "bird_name") %>%
  rename(cluster = "fit$cluster") %>%
  full_join(birds, by = "bird_name")

# get top labels for each cluster
cluster_labels <- kmeans_out %>%
  group_by(cluster) %>%
  arrange(desc(total), .by_group = TRUE) %>%
  slice_head(n = 2) %>%
  mutate(bird = bird_name) %>%
  select(bird_name, bird)

kmeans_out <- kmeans_out %>%
  full_join(cluster_labels) 

kmeans_out$bird[is.na(kmeans_out$bird)]<- ""

```

The next visualilzation section is inspired by <https://www.datanovia.com/en/blog/k-means-clustering-visualization-in-r-step-by-step-guide/>

**Create Data for Plotting**

```{r}
# Dimension reduction using PCA
res.pca <- prcomp(bird_matrix,  scale = TRUE, rank = 4)
# Coordinates of individuals
ind.coord <- as.data.frame(get_pca_ind(res.pca)$coord)
# Add clusters obtained using the K-means algorithm
ind.coord$cluster <- factor(fit$cluster)
# Add Species groups from the original data sett
ind.coord$total_count <- birds$total
ind.coord$bird <- kmeans_out$bird

# Data inspection
#head(ind.coord)

# Percentage of variance explained by dimensions
eigenvalue <- round(get_eigenvalue(res.pca), 1)
variance.percent <- eigenvalue$variance.percent
#head(eigenvalue)
```

**plot**

```{r}
p <- ggscatter(
  ind.coord, x = "Dim.1", y = "Dim.2", 
  color = "cluster", palette = "npg", ellipse = TRUE, ellipse.type = "convex",
   size = 1.5,  legend = "right", ggtheme = theme_bw(),
  xlab = paste0("First Cluster Dimension (variance:", variance.percent[1], "% )" ),
  ylab = paste0("Second Cluster Dimension (variance:", variance.percent[2], "% )" )
) +
  geom_label_repel(aes(label = bird, color = cluster) , force = 5,
                   min.segment.length = 0.1,
                  max.overlaps = 200, fill = alpha(c("white"),0.5) ,
                  show.legend = FALSE) +
  labs(
    title = "Birds of a Feather, a #TidyTuesday plot",
    subtitle = "Common Clusters of Birds seen at Given Austalian Birdbaths",
    caption = "Data from Cleary et al, 2016: https://doi.org/10.1371/journal.pone.0150899 ")

p

# alt text:  A scatter plot showing 4 clusters of Australian birds commonly seen together. Australian Magpies and Noisy Miners are grouped together in an outlier.
```

It's not unlikely that these clusters overlap with the region and season variables, but that's an area for future exploration, and beyond the scope of this \#TidyTuesday exercise.

**Export**

```{r}
ggsave( "TT_Bird_clusters.png",p, dpi = 600)
```
